Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-14 paper/paper.tex
update the style of subsubsection

In the first case of pattern matching, we may have multiple choices of closure for callee @f@.
For each closure in the set, a new frame is constructed and pushed onto the stack.
The code for handling the second case (atomic expressions) is the same as the concrete CESK
machines.

\textbf{Note on Computability.}
Unfortunately, even though other components in the states are finite,
this AAM with an unbounded stack is still not computable.
This is because the unbounded stack can grow to arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

An additional argument @continue@ which has type @Cont@ is also introduced.
Recall that in the disentangled AAM, the second case in @drive_step@
makes a call to @continue@. Here, the reemerging @continue@ is a function
of type @Ans => Ans@. The function @aeval@ does not return: instead, it calls
@continue@ with the result values.

\textbf{Nondeterminism Abstractions.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:

@@

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is a variant implementation of @fold@ in continuation style. But here we have two
continuations, one for the next element in the collection, and one for the eventual result.
This is sometimes referred to as an extended continuation-passing style (ECPS)
that function @k@ is a normal continuation and @m@ is a meta-continuation 
\cite{Danvy:1990:AC:91556.91622}.

\textbf{Evaluation without Caching.}
Since our first version of @aeval@ implements only the nondeterministic evaluation, 
given the defined @nd@ operator, the evaluation function @aeval@ can be easily defined. 
In the case that @e@ is a @let@ expression, we invoke @nd@ twice and perform a nesting 
two-step, depth-first evaluation over possible values of @App(f, ae)@.
The values of branches from a fork point will be accumulated inside of @nd@, and 
then returned to the upper fork point.

@@

Based on the refunctionalized AAM from the last section, we use the same nondeterministic
operator @nd@, but further extend it to a fixed-point cached version in this section.
The fixed-point caching algorithm performs a sound over-approximation of all possible
concrete reachable paths and values, and also prevents non-termination when analyzing 
programs that may diverge.

\textbf{Configuration.}
We do not have the concept of an explicit stack or state, so to represent the cache 
we introduce a configuration @Config@ as a state-like definition which borrows the 
components from state, sans continuations.

@@

\textbf{Cache.}
Recall that the latent assumption of the worklist algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.

@@

The @Cache@ contains two maps @in@ and @out@ which are both mappings from configurations 
to sets of @VS@. The @in@ cache contains mappings from the previous iteration of evaluation, 
and the @out@ cache contains mappings after the current iteration of evaluation. 
Once we have evaluated a term to some values, we update the @out@ cache. In the next 
iteration, we will use the @out@ from the previous iteration as the @in@ cache for the 
current iteration.

\textbf{Cache-Passing Style.}
We can now transform our interpreter into cache-passing style. We begin by changing the 
return type @Ans@ to a case class which contains a @VS@ set and a cache. @Ans@ also implements 
its own accumulating operation @++@ which defines as union on @Set[VS]@ and join on caches.
The join operation of caches is defined as joining its two store components.

@@

In the previous section, we have established a computable pushdown control-flow
analysis through refunctionalization and caching. In this section,
we revisit the pushdown control-flow problem and examine what we have done to
overcome it.

\textbf{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models
the runtime call stack of the analyzed program.
A pushdown control-flow analysis provides an as-exact-as-runtime return-flow
when analyzing a program, but traditional control-flow analysis collapses
the state space into a finite space and thus causes imprecise stack modeling.

@@

In $k$-CFA algorithm or the abstract abstract machine shown in Section~\ref{aam},
the call sites @(id 2)@ and @(id 1)@ share the same return flow,
so the invocation of @(id 2)@ returns to both call-sites @[x (id 1)]@ and @[y (id 2)]@.
Therefore, the returned value @2@ for variable @y@ is also propagated to
variable @x@, causing imprecise analysis results to arise.
This return-flow merging is inevitable, even when increasing the context-sensitivity.
If we use the monovariant analysis (i.e., $0$-CFA), the analysis result would be such 
that @x@ and @y@ point to value set @{1, 2}@, because the algorithm does not distinguish 
that @z@ comes from a different call site.
Under $1$-CFA, the algorithm is able to distinguish that variable @z@ of function
@id@ has two different values at two call sites, so variable @y@
would not be polluted by @1@.
However, variable @x@ still points to value set @{1, 2}@, because two call-sites still 
share the same continuation and $1$-CFA is still unable to separate the return-flows.

\textbf{Call/Return Matching through Refunctionalization and Caching.}
The refunctionalized AAM we obtained has the perfect match on return-flows even though 
we do not have a stack model. The higher-order functions representing continuations 
already connect all the execution in order. In fact, the continuations (call stack) 
of the analyzed language is blended into the call stack of our defining language 
(Scala) through refunctionalization, then the call/return of the analyzed language 
is naturally matched.

@@

To address this problem, we reintroduce two operators @nd@ and @choices@ for handling 
nondeterministic choices in a way that is direct-style but mimics extended 
continuation-passing style.

\textbf{Nondeterminism Operators, Again.}
%Introducing @Cache@ explicitly as the part of arguments of continuation @k@ reveals 
%that cache should be monotonically accumulated to the rest of computation. 
The first operator @nd@ is adapted from the previous one, but the meta-continuation
@m@ has been eliminated here. Now the only continuation @k@ applies on tuple type @(T, Cache)@ 
where the value of type @T@ comes from the collection, and @nd@ performs recursive invocation
on itself over elements in the collection @ts@.

>>>>

In the first case of pattern matching, we may have multiple choices of closure for callee @f@.
For each closure in the set, a new frame is constructed and pushed onto the stack.
The code for handling the second case (atomic expressions) is the same as the concrete CESK machines. \\

\textit{Note on Computability.}
Unfortunately, even though other components in the states are finite,
this AAM with an unbounded stack is still not computable.
This is because the unbounded stack can grow to arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

An additional argument @continue@ which has type @Cont@ is also introduced.
Recall that in the disentangled AAM, the second case in @drive_step@
makes a call to @continue@. Here, the reemerging @continue@ is a function
of type @Ans => Ans@. The function @aeval@ does not return: instead, it calls
@continue@ with the result values. \\

\textit{Nondeterminism Abstractions.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:

@@

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is a variant implementation of @fold@ in continuation style. But here we have two
continuations, one for the next element in the collection, and one for the eventual result.
This is sometimes referred to as an extended continuation-passing style (ECPS)
that function @k@ is a normal continuation and @m@ is a meta-continuation 
\cite{Danvy:1990:AC:91556.91622}. \\

\textit{Evaluation without Caching.}
Since our first version of @aeval@ implements only the nondeterministic evaluation, 
given the defined @nd@ operator, the evaluation function @aeval@ can be easily defined. 
In the case that @e@ is a @let@ expression, we invoke @nd@ twice and perform a nesting 
two-step, depth-first evaluation over possible values of @App(f, ae)@.
The values of branches from a fork point will be accumulated inside of @nd@, and 
then returned to the upper fork point.

@@

Based on the refunctionalized AAM from the last section, we use the same nondeterministic
operator @nd@, but further extend it to a fixed-point cached version in this section.
The fixed-point caching algorithm performs a sound over-approximation of all possible
concrete reachable paths and values, and also prevents non-termination when analyzing 
programs that may diverge. \\

\textit{Configuration.}
We do not have the concept of an explicit stack or state, so to represent the cache 
we introduce a configuration @Config@ as a state-like definition which borrows the 
components from state, sans continuations.

@@

\textit{Cache.}
Recall that the latent assumption of the worklist algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.

@@

The @Cache@ contains two maps @in@ and @out@ which are both mappings from configurations 
to sets of @VS@. The @in@ cache contains mappings from the previous iteration of evaluation, 
and the @out@ cache contains mappings after the current iteration of evaluation. 
Once we have evaluated a term to some values, we update the @out@ cache. In the next 
iteration, we will use the @out@ from the previous iteration as the @in@ cache for the 
current iteration. \\

\textit{Cache-Passing Style.}
We can now transform our interpreter into cache-passing style. We begin by changing the 
return type @Ans@ to a case class which contains a @VS@ set and a cache. @Ans@ also implements 
its own accumulating operation @++@ which defines as union on @Set[VS]@ and join on caches.
The join operation of caches is defined as joining its two store components.

@@

In the previous section, we have established a computable pushdown control-flow
analysis through refunctionalization and caching. In this section,
we revisit the pushdown control-flow problem and examine what we have done to
overcome it. \\

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models
the runtime call stack of the analyzed program.
A pushdown control-flow analysis provides an as-exact-as-runtime return-flow
when analyzing a program, but traditional control-flow analysis collapses
the state space into a finite space and thus causes imprecise stack modeling.

@@

In $k$-CFA algorithm or the abstract abstract machine shown in Section~\ref{aam},
the call sites @(id 2)@ and @(id 1)@ share the same return flow,
so the invocation of @(id 2)@ returns to both call-sites @[x (id 1)]@ and @[y (id 2)]@.
Therefore, the returned value @2@ for variable @y@ is also propagated to
variable @x@, causing imprecise analysis results to arise.
This return-flow merging is inevitable, even when increasing the context-sensitivity.
If we use the monovariant analysis (i.e., $0$-CFA), the analysis result would be such 
that @x@ and @y@ point to value set @{1, 2}@, because the algorithm does not distinguish 
that @z@ comes from a different call site.
Under $1$-CFA, the algorithm is able to distinguish that variable @z@ of function
@id@ has two different values at two call sites, so variable @y@
would not be polluted by @1@.
However, variable @x@ still points to value set @{1, 2}@, because two call-sites still 
share the same continuation and $1$-CFA is still unable to separate the return-flows. \\

\textit{Call/Return Matching through Refunctionalization and Caching.}
The refunctionalized AAM we obtained has the perfect match on return-flows even though 
we do not have a stack model. The higher-order functions representing continuations 
already connect all the execution in order. In fact, the continuations (call stack) 
of the analyzed language is blended into the call stack of our defining language 
(Scala) through refunctionalization, then the call/return of the analyzed language 
is naturally matched.

@@

To address this problem, we reintroduce two operators @nd@ and @choices@ for handling 
nondeterministic choices in a way that is direct-style but mimics extended 
continuation-passing style. \\

\textit{Nondeterminism Operators, Again.}
%Introducing @Cache@ explicitly as the part of arguments of continuation @k@ reveals 
%that cache should be monotonically accumulated to the rest of computation. 
The first operator @nd@ is adapted from the previous one, but the meta-continuation
@m@ has been eliminated here. Now the only continuation @k@ applies on tuple type @(T, Cache)@ 
where the value of type @T@ comes from the collection, and @nd@ performs recursive invocation
on itself over elements in the collection @ts@.
